Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors for COVID-19 in Scotland (REACT-SCOT): a population-based case-control study

### 2020-06-12 casecontrol.Rmd
fixed to install required packages

  - code: IGMM
    address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology.

@@

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true

csl: ../plos.csl
urlcolor: blue
---

@@

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)

@@

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

@@

@@

_Methods_ -- The design was a matched case-control study. Severe cases were all those with a positive nucleic acid test for SARS-CoV-2 in the national database who had entered a critical care unit or died within 28 days of the first positive test.  Ten controls per case matched for sex, age and primary care practice were selected from the population register.  All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.  Rate ratios for severe COVID-19 were estimated by conditional logistic regression.

_Findings_ -- In a logistic regression using the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.  Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r table.agegr.all[5, 3]` for Type 1 diabetes, `r table.agegr.all[6, 3]` for Type 2 diabetes, 
`r table.agegr.all[8, 3]` for ischemic heart disease,
`r table.agegr.all[9, 3]` for other heart disease, 
`r table.agegr.all[10, 3]` for chronic lower respiratory tract disease, 
`r table.agegr.all[11, 3]` for chronic kidney disease, 
`r table.agegr.all[12, 3]` for neurological disease, 
`r table.agegr.all[13, 3]` for chronic liver disease and 
`r table.agegr.all[14, 3]` for immune deficiency or suppression.

@@

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.  Individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` in ECOSS were ascertained.  Using the Community Health Index (CHI) identifier contained in ECOSS (the CHI number is a unique identifier used in all care systems in Scotland) linkage to other datasets was carried out. Hospital admissions from the time of testing were obtained from the RAPID database a daily return of current hospitalisations each day.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers admissions to critical care [comprising adult intensive care units (ICUs), high dependency units (HDUs) and combined ICU / HDU units] across Scotland and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic. Death registrations up to `r gsub("^0", "", format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y"))` were obtained from linkage to the National Register of Scotland.

@@

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one-year age band and registered with the same primary care practice who were alive on the date of the first test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.

## Demographic data
Residence in a care home was ascertained from the CHI database. Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD), an indicator based on postal code.  Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.  For 54% of cases and 28% of controls self-assigned ethnicity, based on the categories used in the Census, had been recorded in Scottish Morbidity Records (SMR).  Cross-tabulation of `r sum.xtabulate` records for which both name classification and SMR records of ethnicity were available showed that the ONOMAP algorithm had sensitivity of `r round(SouthAsian.sensitivity)`% and specificity of `r round(SouthAsian.specificity, 2)`% for classifying South Asian ethnicity, but misclassified most of those who identified as African, Caribbean or Black.

@@

The eight listed conditions were scored based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes. The R script included as supplementary material contains the derivations of these variables from ICD-10 codes and BNF drug codes.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown. Cases of diabetes diagnosed since the last update of the register were identified through discharge codes and drug codes, and assigned to the diabetes type Other/Unknown category.

@@

For the case-control study,  all estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.  Where an exclusion criterion such as having a pre-defined condition was applied to cases  this was also applied to controls as otherwise subsequent association estimates would be incorrect. Where the sample of cases and controls is restricted, this will generate strata that contain no cases but these strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.  With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.  Note that odds ratios in a matched case control study are based on the conditional likelihood and  the unconditional odds ratios calculable from the frequencies of exposure in cases and controls will differ from these and should not be used [@breslow_estimation_1978].
Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios but with less precision than in an unmatched study of the same size [@breslow_estimation_1978].

@@

### Factors derived from hospitalisation and prescribing records
Prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[4, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[4, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[4, 8])` of the cases and `r gsub("\\\\", "", table.agegr[4, 7])` of the controls had at least one listed condition.  The proportion with at least one dispensed prescription was much higher in cases than in controls in each age group. Among those aged under 40 years, `r table.agegr[6, 2]` of the cases and `r table.agegr[6, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.

Over all age groups, `r freqs.all[nrow(freqs.all), 2]` of severe cases and `r freqs.all[nrow(freqs.all), 1]` of controls had at least one of the listed conditions.  As shown in Table \ref{tab:listed.conditions.allages}, all the listed conditions were more frequent in cases than controls.  The rate ratio associated with type 1 diabetes was higher than that for type 2 diabetes.  The rate ratio was `r table.agegr.all[8, 3]` for ischemic heart disease compared to `r table.agegr.all[9, 3]` for the broad category "other heart disease".  In multivariate analysis ischemic heart disease was not independently associated with severity whereas other heart disease remained strongly associated.

Supplementary Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 year age bands combined.  All listed conditions were associated with severe disease in each age band.  In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][5, 3]` for Type 1 diabetes and `r tables.agegr[[1]][6, 3]` for Type 2 diabetes.  The multivariate analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and within each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.  Table \ref{tab:fatal} shows that in each age group the proportion of fatal cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was very low.

## Systematic analysis of diagnoses associated with severe disease
The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, among those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  Note that hospital diagnoses classified under the pregnancy chapter here are derived from admissions with pregnancy related medical conditions to non-obstetric units only, as obstetric returns are not in the SMR01 dataset. In a multivariate analysis the strongest associations were with diagnoses in ICD chapter 2 (neoplasms). Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.  This table is filtered to show only subchapters for which the univariate p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.  This shows that many diagnoses are associated with markedly higher risk of severe COVID-19.  Past hospital diagnoses of infections, pneumonia and acute respiratory diseases were strongly associated with severe COVID-19.  Cardiovascular diagnoses associated with COVID-19 were not limited to heart disease but included also stroke and other circulatory disorders that are not designated as risk conditions.

## Associations of prescribed drugs with severe disease 
As shown in Table \ref{tab:listed.conditions.allages} and supplementary tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} , the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of at least one prescription in the last year.  The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` respectively.  About one third of controls aged over 75 had not encashed a prescription in the previous year.

To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.  In a multivariate analysis of all chapters, most of these associations were weaker.  The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular).  Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).

@@

Although the numbers of cases and controls of non-White ethnicity are small and the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White ethnic groups up to now.  The only non-White ethnic group with any sizeable numbers is the South Asian category and we found no evidence of any elevation in risk in this group compared to Whites.  Reports from England [@the_opensafely_collaborative_opensafely_2020] found elevation in risks for some non-White groups. In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status.  In a study of risk factors for hospitalized disease in the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated but did not fully explain the elevated crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].  The relative socioeconomic position of ethnic groups in Scotland is different to that in England, so it is plausible that the relation of health status to ethnicity will also differ.  For example in the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.  Among the 1.0% who identified as Pakistani or Bangladeshi the proportion living in the most deprived neighbourhoods was not higher than the national average [@walsh_increasingly_2019].  Future work may allow more complete assignment of ethnicity and disaggregation of broad categories based on continent of origin.

@@

Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.  However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were much higher in cases than in controls.  In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.

A striking finding of this study was the strong association of severe COVID-19 with having encashed at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.  Also associated were those in the cardiovascular, nervous system and nutritional and blood chapters.  Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.  These associations are explored in an accompanying paper.

@@

As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.  A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 1)` associated with a 20-year increase in age is far stronger than that associated with common diseases such as Type 2 diabetes and asthma that are listed as conditions associated with high risk.  A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.  It will also exclude many people at high risk because they have multiple risk factors each of small effect.  The only way to optimize risk classification so as to ensure equity with respect to risk is to construct a classifier that uses all available information to assign a risk score.  Our results show that this is possible in principle, though for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem.  In Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.

@@

This case-control study is limited to test-positive cases, excluding deaths with no record of a positive test where COVID-19 was mentioned on the death certificate as an underlying or contributing cause.  Up to 13 May 2020 an additional 1200 such deaths had been reported by the National Records of Scotland.  Future analyses of this study will include sensitivity analyses of the extent to which the results are changed by including these deaths, but linked data are not yet available to us.  Apart from residential care home status, we do not expect most other risk factors to differ markedly between those who died from COVID-19 without being tested and those who died after testing.

@@

knitr::kable(table.demog.append[, 1:5],
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             row.names=FALSE,
             col.names=c("", clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align="lcccc", 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="1.8cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
    group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
    group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
               hline_before=TRUE, latex_gap_space="10pt") %>% 
    kable_styling(latex_options="hold_position")

@@

rownames(table.agegr) <- replace.names(rownames(table.agegr))
knitr::kable(table.agegr[-7, ], # omit non-diabetic category
             escape=FALSE, 
             booktabs=TRUE,
             label="agegr",
             align=rep("c", 8), 
             caption="Frequencies of risk factors in cases and controls, by age group") %>%
    column_spec(column=1, width="4cm") %>%
    column_spec(column=2:9, width="2cm") %>%
    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
                       "60-74 years"=2, "75+ years"=2)) %>%
    landscape() %>%
    kable_styling(latex_options="hold_position")

@@

>>>>

  - code: IGMM
    address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology.

@@

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ./covid.bib

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true

always_allow_html: true
csl: ./plos.csl
urlcolor: blue
---

@@

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

@@

options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

@@

table.ethnicsmr.missing <- paste.colpercent(with(cc.severe, table(is.na(ethnic5.smr), CASE)))

table.notlisted.scrip <- paste.colpercent(with(cc.severe[notlisted, ], table(scrip.any, CASE)))
table.notlisted.diag <- paste.colpercent(with(cc.severe[notlisted, ], table(diag.any, CASE)))
table.notlisted.scripordiag <- paste.colpercent(with(cc.severe[notlisted, ], table(scripordiag, CASE)))

table.lt60.notlisted.scripordiag <- paste.colpercent(with(cc.severe[cc.severe$AGE < 60 & notlisted, ], table(scripordiag, CASE)))

@@

_Methods_ -- The design was a matched case-control study. Severe cases were all those with a positive nucleic acid test for SARS-CoV-2 in the national database who had entered a critical care unit or died within 28 days of the first positive test.  Seven controls per case matched for sex, age and primary care practice were selected from the population register.  All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.  Rate ratios for severe COVID-19 were estimated by conditional logistic regression.

_Findings_ -- There were `r length(which(cc.severe$CASE==1))` severe cases. In a logistic regression using the age-sex distribution of the national population, the odds ratios for severe disease were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.  Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r table.agegr.all[5, 3]` for Type 1 diabetes, `r table.agegr.all[6, 3]` for Type 2 diabetes, 
`r table.agegr.all[8, 3]` for ischemic heart disease,
`r table.agegr.all[9, 3]` for other heart disease, 
`r table.agegr.all[10, 3]` for chronic lower respiratory tract disease, 
`r table.agegr.all[11, 3]` for chronic kidney disease, 
`r table.agegr.all[12, 3]` for neurological disease, 
`r table.agegr.all[13, 3]` for chronic liver disease and 
`r table.agegr.all[14, 3]` for immune deficiency or suppression.

@@

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` in ECOSS were ascertained for this study.  Using the Community Health Index (CHI) identifier contained in ECOSS (the CHI number is a unique identifier used in all care systems in Scotland) linkage to other datasets was carried out. Hospital admissions from the time of testing were obtained from the RAPID database a daily return of current hospitalisations each day.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers admissions to critical care [comprising adult intensive care units (ICUs), high dependency units (HDUs) and combined ICU / HDU units] across Scotland and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic. Death registrations up to `r gsub("^0", "", format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y"))` were obtained from linkage to the National Register of Scotland.

@@

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one-year age band and registered with the same primary care practice who were alive on the date of the first test in the case using the Community Health Index (CHI) database. After removing embarkations there were seven controls per case. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.

## Demographic data
Residence in a care home was ascertained from the CHI database. Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD), an indicator based on postal code.  Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.  For `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls and `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases self-assigned ethnicity, based on the categories used in the Census, had been recorded in Scottish Morbidity Records (SMR).  Cross-tabulation of `r sum.xtabulate` records for which both name classification and SMR records of ethnicity were available showed that the ONOMAP algorithm had sensitivity of `r round(SouthAsian.sensitivity)`% and specificity of `r round(SouthAsian.specificity, 2)`% for classifying South Asian ethnicity, but misclassified most of those who identified as African, Caribbean or Black.

@@

The eight listed conditions were scored based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes. The R script included as supplementary material contains the derivations of these variables from ICD-10 codes and BNF drug codes.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown.

@@

For the case-control study, all estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.  Restriction of cases and controls, for instance to exclude those with any listed condition, may generate strata that do not contain at least one case and at least one control, but these strata are ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.  With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.  Note that odds ratios in a matched case control study are based on the conditional likelihood and the unconditional odds ratios calculable from the frequencies of exposure in cases and controls will differ from these and should not be used [@breslow_estimation_1978].
Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios but with less precision than in an unmatched study of the same size [@breslow_estimation_1978].

@@

### Factors derived from hospitalisation and prescribing records
Prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.  `r gsub("\\\\", "", table.agegr[4, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[4, 1])` of the controls.  In those aged 75+ years `r gsub("\\\\", "", table.agegr[4, 8])` of the cases and `r gsub("\\\\", "", table.agegr[4, 7])` of the controls had at least one listed condition.  Among those aged under 40 years, `r table.agegr[6, 2]` of the cases and `r table.agegr[6, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.  Differences in prescription rates  between cases an controls narrowed with increasing age.

Over all age groups, `r freqs.all[nrow(freqs.all), 2]` of severe cases and `r freqs.all[nrow(freqs.all), 1]` of controls had at least one of the listed conditions.  As shown in Table \ref{tab:listed.conditions.allages}, all the listed conditions were more frequent in cases than controls except for immune conditions in the 75+ age group.  The rate ratio associated with type 1 diabetes was higher than that for type 2 diabetes.  The rate ratio was `r table.agegr.all[8, 3]` for ischemic heart disease compared to `r table.agegr.all[9, 3]` for the broad category "other heart disease".  In multivariate analysis ischemic heart disease was not independently associated with severity whereas other heart disease remained strongly associated. In those without a listed condition `r table.notlisted.scripordiag[2, 2]` of the cases and `r table.notlisted.scripordiag[2, 1]` of the controls had either a recent admission or a prescription.  In those aged under 60 years without a listed condition, `r table.lt60.notlisted.scripordiag[2, 2]` of the cases and `r table.lt60.notlisted.scripordiag[2, 1]` had either a recent admission or a prescription.

Supplementary Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 year age bands combined.  All listed conditions were associated with severe disease in each age band.  In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][5, 3]` for Type 1 diabetes and `r tables.agegr[[1]][6, 3]` for Type 2 diabetes.  The multivariate analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and in each age group  group  any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions. Dispensing of any prescription in the past year was associated with severe disease  in multivariate analyses in the two younger age bands.  Table \ref{tab:fatal} shows that in each age group the proportion of fatal cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was very low.

In a sensitivity analysis in which we also included deaths registered with mention of COVID-19 on the death certificate in the definition of severe cases (Table \ref{tab:agegr.withNRS}) the same pattern of differences in prior admissions and prescribing history and in listed conditions between cases and controls was found, with the difference in residential are home status being somewhat greater. Such deaths were not included in our primary outcome definition as misclassification rates for COVID assignation as cause are unknown.

## Systematic analysis of diagnoses associated with severe disease
The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, among those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 7 (eye) Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  Note that hospital diagnoses classified under the pregnancy chapter here are derived from admissions with pregnancy related medical conditions to non-obstetric units only, as obstetric returns are not in the SMR01 dataset. In a multivariate analysis the most significant association was with diagnoses in ICD chapter 2 (neoplasms). Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.  This table is filtered to show only subchapters for which the univariate p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.  This shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.  <!-- check why cerebrovascular disease does not come out here -->

## Associations of prescribed drugs with severe disease 
As shown in Table \ref{tab:listed.conditions.allages} and supplementary tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} ,  encashment of at least one prescription in the last year was associated with severe disease.  The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` respectively.

To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.  In a multivariate analysis of all chapters, most of these associations were weaker.  The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and chapters 4 (central nervous system).  Other chapters associated with severe disease were 2 (cardiovascular),  5 (infections), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).

@@

Although the numbers of cases and controls of non-White ethnicity are small and the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White ethnic groups up to now.  The only non-White ethnic group with any sizeable numbers is the South Asian category and we found no clear evidence of any elevation in risk in this group compared to Whites.  Reports from England [@the_opensafely_collaborative_opensafely_2020] found elevation in risks for some non-White groups. In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status.  In a study of risk factors for hospitalized disease in the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated but did not fully explain the elevated crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].  The relative socioeconomic position of ethnic groups in Scotland is different to that in England, so it is plausible that the relation of health status to ethnicity will also differ.  For example in the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.  Among the 1.0% who identified as Pakistani or Bangladeshi the proportion living in the most deprived neighbourhoods was not higher than the national average [@walsh_increasingly_2019].  Future work may allow more complete assignment of ethnicity and disaggregation of broad categories based on continent of origin.

@@

Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.  However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were higher in cases than in controls.  In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.

An important finding of this study was the strong association of severe COVID-19 with having encashed at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.  Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.  Also associated were those in the  nervous system,  cardiovascular, and nutritional and blood chapters.  Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.  These associations are explored in an accompanying paper.

@@

As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.  A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 1)` associated with a 20-year increase in age is stronger than that associated with common diseases such as Type 2 diabetes and asthma that are listed as conditions associated with high risk.  A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.  It will also exclude many people at high risk because they have multiple risk factors each of small effect.  The only way to optimize risk classification so as to ensure equity with respect to risk is to construct a classifier that uses all available information to assign a risk score.  Our results show that this is possible in principle, though for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem.  In Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.

@@

@@

knitr::kable(table.demog.append[, 1:5],
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             row.names=FALSE,
             col.names=c("", clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align="lcccc", 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="1.8cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    kableExtra::group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
    kableExtra::group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
    kableExtra::group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
               hline_before=TRUE, latex_gap_space="10pt") %>% 
    kable_styling(latex_options="hold_position")

@@

rownames(table.agegr) <- replace.names(rownames(table.agegr))
table.agegr.print <- table.agegr[c(1:4, 6, 8:nrow(table.agegr)), ] 
knitr::kable(table.agegr.print, 
             escape=FALSE, 
             booktabs=TRUE,
             label="agegr",
			 linesep=c(rep('', 4), '\\addlinespace', rep('', 9)), 
             align=rep("c", 8), 
             caption="Frequencies of risk factors in cases and controls, by age group") %>%
    column_spec(column=1, width="4cm") %>%
    column_spec(column=2:3, width="1.6cm") %>%
    column_spec(column=4:7, width="1.8cm") %>%
    column_spec(column=8:9, width="2cm") %>%
    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
                       "60-74 years"=2, "75+ years"=2)) %>%
    landscape() %>%
    kable_styling(latex_options="hold_position")

@@

\newpage
\beginsupplement
